---
title: "Hedge Fund Clusters"
author: "James Walsh"
date: "5/7/2022"
output: html_document
---

Cluster analysis is an unsupervised learning methodology within machine learning. The approach is straight-forward, it attempts to look for clusters in data based on a definition of distance between key characteristics. In this analysis the Euclidean distance is used, which is the squared distance between normalized characteristics. The characteristics are shown in the table below.

There are a number of reasons why understanding the similarity and differences between strategies may be of interest.

1.  When building portfolios there is often a practical limitation on the number of funds or investments that can be incorporated in a portfolio. By understanding which strategies are indeed similar we can narrow the number of possible strategies.

2.  Similarities can change over time.

3.  Clustering takes into account not only the systematic characteristics of a strategy, but also the idiosyncratic returns.

Another benefit of this approach is it doesn't require require that a number of clusters is pre-specified, rather they are revealed by the data depending on the characteristics in the data that are deemed important.

```{r setup, include=FALSE}
library(cluster)
library(tidyverse)
library(factoextra)
library(dendextend)
library(moments)
library(dplyr)
library(NbClust)
library(pheatmap)
library(tidyquant)
library(PortfolioAnalytics)
library(PerformanceAnalytics)

load("HedgeRS_EW.RData")
attach(HedgeRSReturns.t)

end_date<-last(HedgeRSReturns.t$date)
start_date<-first(HedgeRSReturns.t$date)
```

```{r functions, echo=FALSE}

#This calculates stress beta using APP_Equity
stressbeta<-function(x){
    fit<-lm(x~APP_Equity,data = stress.t)
    summary(fit)$coefficients[2]
}
##This calculates the beta
beta<-function(x){
  fit<-lm(x~APP_Equity,data=Hedge_data.t)
  x<-summary(fit)$coefficients[2]
}


```

### Statistics used in cluster analysis

Time period: December 2003 - July 2021 Data: HedgeRS Equally Weighted

```{r dataprep,echo=FALSE}
c<-length(HedgeRSReturns.t)-9
columns<-c(2:c)

#reduce to desired time period

rows<-c(1:nrow(HedgeRSReturns.t))
Hedge_data.t<-select(HedgeRSReturns.t,-HedgeRS_EW,-Multi_EW)
Hedge_data.t<-Hedge_data.t[rows,]

#Create states
z <- Hedge_data.t$VIX #load VIX into z
score <- (z-mean(z, na.rm = TRUE))/sd(z, na.rm = TRUE) #calculate z score

Hedge_data.t<-cbind(Hedge_data.t,score) #add z score to the data.table

Hedge_data.t$state<-"Central" #create new column based on score - named as state
Hedge_data.t$state[score>=1]<-"High"
Hedge_data.t$state[score<=1]<-"Low"


##Create stressed subset
stress.t<-Hedge_data.t%>%
filter(state=="High")
#Calculate the stats

HF_beta<-sapply(Hedge_data.t[,columns],beta)
HF_sd<-sapply(Hedge_data.t[,columns],sd)*(12^0.5)
HF_mean<-sapply(Hedge_data.t[,columns],mean)*12
HF_skew<-sapply(Hedge_data.t[,columns],skewness)
HF_kurt<-sapply(Hedge_data.t[,columns],kurtosis)
Sharpe<-(HF_mean-(mean(Hedge_data.t$Cash)*12))/HF_sd
Treynor<-(HF_mean-(mean(Hedge_data.t$Cash)*12))/HF_beta
Stress<-sapply(stress.t[,columns],stressbeta)
```

```{r cluster_data,echo=FALSE}
#df<-cbind(HF_sd,HF_mean,HF_skew,HF_kurt,HF_beta,Sharpe,Treynor,Stress)
df_base<-cbind(HF_mean,HF_sd,HF_skew,HF_kurt,Stress)
#df_base<-cbind(HF_mean,HF_sd)

HF_names<-c("CTA","EM GM","GAA","Global Macro","Insurance","L/S Asia Pac",
            "LS Emerging","LS Europe","LS Global","LS Japan","LS US",
            "Activist","Distressed","EM FI","Risk Arb","CB Arb",
            "RV Credit","RV FI","FEMN","QEMN","Stat Arb","Structured Credit",
            "Multi Event","Multi RV","Multi Diversifited","Multi Opportunistic")

row.names(df_base)<-HF_names
df_base_round<-round(df_base,3)
df_base_round
df<-na.omit(df_base)
df<-scale(df)

```

### Euclidian Distance

The Euclidian Distance is the squared distance from each other based on the above characteristics. A number of close to 0 implies they are similar, and a high number suggests they are very different.

```{r distance, echo=FALSE}
#distance1<-get_dist(df,method="pearson")
#fviz_dist(distance1,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))
#distance2<-get_dist(df,method = "kendall")
#fviz_dist(distance2,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))
#distance3<-get_dist(df,method = "spearman")
#fviz_dist(distance3,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))
#distance4<-get_dist(df,method = "manhattan")
#fviz_dist(distance4,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))

set.seed(1)
distance<-get_dist(df,method="euclidean")
fviz_dist(distance,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))

```

### Cluster illustration

These charts show how strategies would be clustered based on 3 to 8 clusters. The x and y values are derived from PCA, or Principle Component Analysis, of all the characteristcs, and show the percentage of variation that can be explained by these first two components.

```{r create_clusters,echo=FALSE,message=FALSE,warning=FALSE}

k2<-kmeans(df,centers = 3,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,3,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 4,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,4,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 5,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,5,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 6,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,6,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 7,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,7,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 8,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,8,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 9,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,9,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 10,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,10,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 11,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,11,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")

k2<-kmeans(df,centers = 12,nstart = 25)
fviz_cluster(k2,data=df,main="K means")

k2<-pam(df,12,nstart = 25)
fviz_cluster(k2,data=df, main="Clustering around Medoids")


```

These charts show specific characteristics of the strategies with the clusters identified by colour

```{r clusters_by_characteristics, echo=FALSE,message=FALSE}
k2<-kmeans(df,8,nstart = 25)
fviz_cluster(k2,data=df)

k2_results<-aggregate(df_base,by=list(clusters=k2$cluster),mean)
k2_results

k2<-pam(df,k = 8,nstart = 25)
fviz_cluster(k2,data=df)

k2_results<-aggregate(df_base,by=list(clusters=k2$cluster),mean)
k2_results

k2$clustering

results<-cbind(df_base,k2$clustering)

results


df %>%
  as_tibble() %>%
  mutate(cluster = k2$cluster,
         state = row.names(df)) %>%
  ggplot(aes(HF_sd, HF_mean, color = factor(cluster), label = state,)) +
  ggtitle("Strategies plotted by cluster on risk v return axis")+
  labs(x="Standard Deviation",y="Return")+
  geom_text()

df %>%
  as_tibble() %>%
  mutate(cluster = k2$cluster,
         state = row.names(df)) %>%
  ggplot(aes(Sharpe, HF_beta, color = factor(cluster), label = state)) +
   ggtitle("Strategies plotted by cluster on Sharpe v Equity beta axis")+
  labs(x="Sharpe Ratio",y="Equity beta")+
  geom_text()

df %>%
  as_tibble() %>%
  mutate(cluster = k2$cluster,
         state = row.names(df)) %>%
  ggplot(aes(HF_mean, Stress, color = factor(cluster), label = state)) +
   ggtitle("Strategies plotted by cluster on Return v Stress Beta axis")+
  labs(x="Return",y="Stress Beta (beta on >=1sd VIX")+
  geom_text()

k2<-kmeans(df,3,nstart = 25)
fviz_cluster(k2,data=df,
             palette=c("#2E9FDF","#00AFBB","#E7B800","#FC4E07"),
             ellipse.type="euclid",
             star.plot=TRUE,
             repel=TRUE,
             ggtheme=theme_minimal()
)

k2<-kmeans(df,4,nstart = 25)
fviz_cluster(k2,data=df,
             palette=c("#2E9FDF","#00AFBB","#E7B800","#FC4E07"),
             ellipse.type="euclid",
             star.plot=TRUE,
             repel=TRUE,
             ggtheme=theme_minimal()
)

k2<-pam(df,7,nstart = 25)
fviz_cluster(k2,data=df,
             palette=c("#1B9E77","#D95F02","#7570B3","#E7298A","#66A61E","#E6AB02","#A6761D"),
             ellipse.type="euclid",
             star.plot=TRUE,
             repel=TRUE,
             ggtheme=theme_minimal()
)


```

### Optimal number of clusters - various methods

```{r k_mean_optimal_clusters,echo=FALSE}

set.seed(123)

wss<-function(k) {
  kmeans(df,k,nstart=10)$tot.withinss
}

k.values<-1:20

wss_values<-map_dbl(k.values,wss)

plot(k.values,wss_values,
     type="b",pch=19,frame=FALSE,
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

fviz_nbclust(df,kmeans,method="wss")

fviz_nbclust(df,kmeans,method="silhouette")

gap_stat <- clusGap(df, FUN = kmeans, nstart = 12, d.power=2,
                    K.max = 20, B = 200)

fviz_gap_stat(gap_stat)
```
```{r pam_optimal_clusters,echo=FALSE}

fviz_nbclust(df,kmeans,method = "silhouette")+ggtitle("k-means clusters")+
  theme_classic()
fviz_nbclust(df,pam,method = "silhouette")+ggtitle("Medoids clusters")+
  theme_classic()

fviz_nbclust(df,kmeans,method = "wss")+ggtitle("k-means clusters")+
  theme_classic()
fviz_nbclust(df,pam,method = "wss")+ggtitle("Medoids clusters")+
  theme_classic()


fviz_nbclust(df,kmeans,nstart=25,method = c("gap_stat"),nboot=100)+
  ggtitle("k-means")+theme_classic()
fviz_nbclust(df,pam,nstart=25,method = c("gap_stat"),nboot=100)+
  ggtitle("Medoids clusters")+theme_classic()

nb<-NbClust(df,distance="euclidean",min.nc=2,max.nc=12,method="centroid")
fviz_nbclust(nb)+ggtitle("centroids")

nb<-NbClust(df,distance="euclidean",min.nc=2,max.nc=12,method="kmeans")
fviz_nbclust(nb)+ggtitle("kmeans")
```

## Hierarchical Clustering

```{r hierarchical_clustering, echo=FALSE}

clusters<-hclust(distance)
plot(clusters)

clusters<-agnes(distance,method = "complete")
plot(clusters)

# methods to assess
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(df, method = x)$ac
}

map_dbl(m, ac)

clusters<-agnes(distance,method = "ward")
plot(clusters)


# compute divisive hierarchical clustering
hc4 <- diana(df)

# Divise coefficient; amount of clustering structure found
hc4$dc
## [1] 0.8514345

# plot dendrogram
pltree(hc4, cex = 0.6, hang = -1, main = "Dendrogram of diana")

# Ward's method
hc5 <- hclust(distance, method = "ward.D2" )

# Cut tree into 4 groups
sub_grp <- cutree(hc5, k = 6)

# Number of members in each cluster
table(sub_grp)

# Compute distance matrix
res.dist <- dist(df, method = "euclidean")

# Compute 2 hierarchical clusterings
hc1 <- hclust(res.dist, method = "complete")
hc2 <- hclust(res.dist, method = "ward.D2")

# Create two dendrograms
dend1 <- as.dendrogram (hc1)
dend2 <- as.dendrogram (hc2)

tanglegram(dend1, dend2)

dend_list <- dendlist(dend1, dend2)

tanglegram(dend1, dend2,
  highlight_distinct_edges = FALSE, # Turn-off dashed lines
  common_subtrees_color_lines = FALSE, # Turn-off line colors
  common_subtrees_color_branches = TRUE, # Color common branches 
  main = paste("entanglement =", round(entanglement(dend_list), 2))
  )

fviz_dend(hc1,cex=0.5,k=8,k_colors="jco",type="circular")

```


```{r correlation_clustering}
cor_matrix<-cor(HedgeRSReturns.t[,2:29])
head(cor_matrix)
cor_matrix_dist<-as.dist(cor_matrix)
cor_matrix_dist
fviz_dist(cor_matrix_dist,gradient=list(low="#00AFBB",mid="white",high="#FC4E07"))

df.corr<-scale(cor_matrix_dist)

fviz_nbclust(df.corr,kmeans,method="wss")

fviz_nbclust(df.corr,kmeans,method="silhouette")

gap_stat <- clusGap(df.corr , FUN = kmeans, nstart = 12, d.power=2,
                    K.max = 20, B = 200)

fviz_gap_stat(gap_stat)

k2<-kmeans(df.corr,centers = 3,nstart = 25)
fviz_cluster(k2,data=df.corr,main="K means")

k2<-pam(df.corr,3,nstart = 25)
fviz_cluster(k2,data=df.corr, main="Clustering around Medoids")



```



```{r portfolios}
funds<-colnames(HedgeRSReturns.t[2:29])
HedgeRSReturns.ts<-ts(HedgeRSReturns.t[2:29])

return_target<-0.06

init <- portfolio.spec(assets=funds) #include insurance
init <- add.constraint(portfolio=init,type="full_investment")
#init <- add.constraint(portfolio = init, type="return", return_target=return_target/12)

min_weights=c(rep(0,length(funds)))
max_weights=c(rep(1,length(funds)))
init <- add.constraint(portfolio=init, type="box", min=min_weights, max=max_weights)
minvar <- add.objective(portfolio=init,type="return",name="mean")
minvar.mv<-add.objective(portfolio=init,type="risk",name="StdDev")
minvar.sd <- add.objective(portfolio=init, type="risk", name="StdDev")
minvar.ed <- add.objective(portfolio=init, type="risk", name="ETL",arguments=list(p=0.95),enabled=TRUE)
minvar.sd.30 <- add.objective(portfolio=init, type="risk_budget", name="StdDev",max_prisk=0.2)
minvar.ed.30 <- add.objective(portfolio=init, type="risk_budget", name="ETL",p=0.95,max_prisk=0.2,enabled=TRUE)
minvar.equal.risk<-add.objective(portfolio=init,type="risk_budget",name="ETL",p=0.95,min_concentration=TRUE,enabled=TRUE)

opt_minvar.sd<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.sd,optimize_method = "ROI",trace = TRUE)
opt_minvar.sd
opt_minvar.mv<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.mv,optimize_method = "ROI",trace = TRUE)
opt_minvar.mv
opt_minvar.ed<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.ed,optimize_method = "ROI",trace = TRUE)
opt_minvar.ed
opt_minvar.sd.30<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.sd.30,optimize_method = "ROI",trace = TRUE)
opt_minvar.sd.30
opt_minvar.ed.30<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.ed.30,optimize_method = "ROI",trace = TRUE)
opt_minvar.ed.30
opt_minvar.equal.risk<-optimize.portfolio(R=HedgeRSReturns.ts,portfolio = minvar.equal.risk,optimize_method = "ROI",trace = TRUE)
opt_minvar.equal.risk

cluster1<-HedgeRSReturns.t$CTA_EW
cluster2<-cbind(HedgeRSReturns.t$GAA_EW,HedgeRSReturns.t$QEMN_EW,HedgeRSReturns.t$Global_Macro_EW,HedgeRSReturns.t$Stat_Arb_EW,
                HedgeRSReturns.t$Risk_Arb_EW,HedgeRSReturns.t$FEMN_EW)
colnames(cluster2)<-c()
cluster3<-cbind(HedgeRSReturns.t$LS_Europe_EW,HedgeRSReturns.t$RV_Credit_EW,HedgeRSReturns.t$Multi_Opp_EW,HedgeRSReturns.t$Multi_Event_EW,
                HedgeRSReturns.t$LS_Japan_EW)


funds<-colnames(cluster2)

cluster2<-Hedge_data.t %>%
  select(GAA_EW,QEMN_EW,Global_Macro_EW,Stat_Arb_EW,Risk_Arb_EW,FEMN_EW)


```

